Persistence in functional topology and a correction to a theorem of Morse

### 2021-09-13 sec/surfaces.tex
Sent to CCM

\section{Minimal surfaces and persistent homology} \label{s:surfaces}

@@

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be a $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and such that there are three distinct points $\alpha_i \in [0,2\pi)$ with $\varphi(\alpha_i)=\alpha_i$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined by
\begin{equation*}
A_g(\varphi) = \frac{1}{16} \int_0^{2\pi} \int_0^{2\pi} \sin \left( \frac{\alpha-\beta}{2} \right)^{-2} \! \lVert g(\varphi(\alpha)) - g(\varphi(\beta)) \rVert_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
Let $\Omega_g = \{\varphi \in \widetilde\Omega \mid A_g(\varphi) < \infty\}$.
Douglas proved that, for a large class of curves $g$, the set $\Omega_g$ is non-empty and the sublevel sets of $A_g$ are compact.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ has a unique global minimizer.
Using methods from harmonic analysis, Douglas then showed that this minimizing critical point of $A_g$ corresponds to a solution of Plateau's Problem.

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, called a minimal surface, and use this correspondence to prove the following result \cite[p.~447]{Morse.1939}, also reviewed as Theorem II.6.10 in \cite{Struwe.1988}.

\begin{thm}[Unstable Minimal Surface Theorem]
	If $\Omega_g$ contains two distinct solutions to Plateau's Problem then it also contains an unstable minimal surface, more precisely, a critical point of $A_g$ of index 1.
\end{thm}

@@

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, in order to obtain \mbox{q-tameness}, Morse assumed slightly varying forms of local connectivity on the resulting sublevel set filtrations.
In particular, in their applications to minimal surface theory, Morse and Tompkins used the following condition:
\begin{displaycquote}[p.~431]{Morse.1940}
	The space $M$ is said to be \emph{locally $F$-connected} of order $r$ at $p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Here, $c = F(p)$.
Using similar language to the one used in \cref{s:connectivity}, this property is equivalent to the following notion applicable to general topological spaces.

\begin{defi}
	The sublevel set filtration of a function $f \colon M \to \R$ is said to be \emph{weakly homotopically locally connected} or \emph{weakly $\piLC$}	if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ induces trivial maps on homotopy groups.
\end{defi}

Morse then goes on to state that as a consequence of local $F$-connectivity the persistent \v{C}ech homology of this sublevel set filtration is q-tame.
In the original it reads:
\begin{displaycquote}[Theorem 6.3, p.~432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, the above claim does not hold in general, as we will show next by providing an explicit counterexample.

@@

Clearly, being weakly $\LC$ implies being weakly $\piLC$ and, if the homology $\H$ takes finite dimensional values on one point spaces, also weakly $\HLC$.
We will now show that not even weakly $\LC$ is sufficient to ensure the q-tameness of compact sublevel set filtrations that are induced by non-continuous functions.
In particular, our construction will be used to invalidate Morse's claim quoted above.

@@

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 60]
	\draw[thick] (-.1,-.001) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05);
	\foreach \i in {1,...,100}{
		\draw[line width=0.1/\i mm] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
\end{figure}

@@

	For $x$ not the origin we have $f(x) = 1$ and choose $s \in (1,t)$ arbitrarily, so that $f_{\leq s} = f_{\leq t} = \HE$.
	Note that since $x$ is not the origin, there is a unique $d$-sphere in $\HE$ that contains $x$.
	Clearly, we may choose $\delta > 0$ so small that $D_{\delta}(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a topological disk contained in this sphere and contained in $V$.
	The disk $D_\delta(x)$ can be contracted to $\{x\}$ in $V$, so choosing $U = D_{\delta}(x)$, we obtain that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to the constant map with value $x$.

@@

Both singular and \v{C}ech homology of the $d$-dimensional Hawaiian earring is infinite dimensional in degree $d$, so they satisfy the assumptions of \cref{t:counterexample}.
For singular homology this is proven in \cite{Barratt.1962}.
For \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6.\@ and X.3.1.]{Eilenberg.1952} as follows.
Define
\begin{align*}
\HE_k &=
\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
\begin{equation*}
\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
\end{equation*}
which is infinite-dimensional over $\F$.

\begin{cor} \label{c:counterexample}
	The function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a weakly $\LC$ compact sublevel set filtration that is not q-tame with respect to singular and \v{C}ech homology.
\end{cor}

>>>>

%!TEX root = ../func_top.tex

\section{Persistent homology and the minimal surface problem} \label{s:surfaces}

@@

Morse and Tompkins considered the following setting introduced by Douglas.
Let $g \colon \R \to \R^n$ be a $2\pi$-periodic function representing a simple closed curve such that $g$ is differentiable with Lipschitz derivative.
Let $\widetilde{\Omega}$ be the space of continuous functions $\varphi \colon \R \to \R$ with $\varphi((t)+2\pi) = \varphi(t) + 2\pi$ for all $t$ and $\varphi(\alpha_i)=\alpha_i$ for three fixed distinct points $\alpha_i \in [0,2\pi)$.
The \emph{Douglas functional} on $\widetilde \Omega$ associated to the curve $g$ is defined by
\begin{equation*}
A_g(\varphi) = \frac{1}{16 \pi} \int_0^{2\pi} \int_0^{2\pi}  \left\| \frac{g(\varphi(\alpha)) - g(\varphi(\beta))}{\sin \frac{\alpha-\beta}{2}} \right\|_2^2 \ \mathrm{d}\alpha \ \mathrm{d}\beta.
\end{equation*}
Let $\Omega_g = \{\varphi \in \widetilde\Omega \mid A_g(\varphi) < \infty\}$.
Douglas proved that, for a large class of curves $g$, the set $\Omega_g$ is non-empty and the sublevel sets of $A_g$ are compact.
Since $A_g$ is bounded below by $0$, this implies that $A_g$ attains a global minimimum.
Douglas then showed that this minimizing critical point of $A_g$ corresponds to a solution of Plateau's Problem, a surface homeomorphic to a disk with boundary $g$ and minimum area.

@@

Morse and Tompkins prove that each homotopically critical point of the functional $A_g$ corresponds to a critical point of the area functional, called a minimal surface \cite[p.~464]{Morse.1939}, and use this correspondence to prove the following result \cite[p.~472]{Morse.1939}, also reviewed in \cite[Theorem II.6.10]{Struwe.1988}.

\begin{thm}[Unstable Minimal Surface Theorem]
	If $\Omega_g$ contains two distinct solutions of Plateau's Problem contained in disjoint critical sets of $A_g$, then it also contains an unstable minimal surface, more precisely, a homotopically critical point of $A_g$ of non-minimum type.
\end{thm}
Moreover, they give an explicit example of a curve $g$ satisfying these assumptions, proving the existence of an unstable minimal surface.

@@

We have seen that these inequalities require q-tameness, so in order to apply them and deduce their theorem, Morse and Tompkins need to prove the \mbox{q-tameness} of $(\Omega_g, A_g)$.
Throughout his work on functional topology, in order to obtain \mbox{q-tameness}, Morse assumed slightly varying forms of local connectivity on the resulting sublevel set filtrations.
In particular, in their applications to minimal surface theory, Morse and Tompkins used the following condition:
\begin{displaycquote}[p.~431]{Morse.1940}
	Let $p$ be a point of $M$ at which $F(p)=c$.
	The space $M$ is said to be \emph{locally $F$-connected} of order $r$ at~$p$ if corresponding to each positive constant $e$ there exists a positive constant $\delta$ such that each singular $r$-sphere on the $\delta$-neighborhood of $p$ and on $F_{c+\delta}$ bounds an $(r+1)$-cell of norm $e$ on $F_{c+e}$.
\end{displaycquote}
Using similar language to the one used in \cref{s:connectivity}, the property of local $F$-connectedness of all orders is equivalent to the following notion applicable to general topological spaces.

\begin{defi}
	The sublevel set filtration of a function $f \colon X \to \R$ is said to be \emph{weakly homotopically locally connected}, or \emph{weakly $\piLC$}, if for any $x \in X$, $V$ a neighborhood of $x$, and any index $t > f(x)$, there is an index $s$ with $f(x) < s < t$ and a neighborhood $U$ of $x$ with $U \subseteq V$ such that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ induces trivial maps on homotopy groups.
\end{defi}

Morse then goes on to claim that the persistent \v{C}ech homology of this sublevel set filtration is q-tame, provided that $F$ is bounded from below and satisfies the assumptions of local $F$-connectivity and compactness of sublevel sets.
%, and a further condition called $F$-regularity, which is trivially satisfied if the domain of $F$ is compact.
In the original (where the function is assumed to take values in $[0,1)$ the claim reads:
\begin{displaycquote}[Theorem 6.3, p.~432]{Morse.1940}
	Let $a$ and $c$ be positive constants such that $a < c < 1$.
	The $k^{\mathrm{th}}$ connectivity $R^k(a,c)$ of $F_a$ on $F_c$ is finite.
\end{displaycquote}
Morse does not prove this statement in the given reference, but rather refers to \cite[Theorem~6.1]{Morse.1938}.
Unfortunately, the above claim does not hold in general, as we will show next by providing an explicit counterexample.
We start by introducing a stronger version of weak local connectedness, which will still be satisfied by our counterexample.

@@

Clearly, being weakly $\LC$ implies being weakly $\piLC$ and, if the homology $\H$ takes finite dimensional values on one-point spaces, also weakly $\HLC$.
We will now show that not even the weak $\LC$ condition is sufficient to ensure the q-tameness of compact sublevel set filtrations that are induced by non-continuous functions.
In particular, our construction will be used to invalidate Morse's claim quoted above.

@@

\begin{figure}[t]
	\centering
	\begin{tikzpicture}[scale = 60]
	\draw[thick] (-.1,-.001) rectangle (.1,.05);
	\clip (-.1,0) rectangle (.1,.05);
	\foreach \i in {1,...,100}{
		\draw[line width=0.4/\i^0.5 pt] (0, 1/\i^2) circle (1/\i^2);
	}
	\end{tikzpicture}
	\caption{A closeup of the Hawaiian earring $\mathbb{H}^1$.}
\end{figure}

@@

	For $x$ not the origin we have $f(x) = 1$ and choose $s \in (1,t)$ arbitrarily, so that $f_{\leq s} = f_{\leq t} = \HE$.
	Note that since $x$ is not the origin, there is a unique $d$-sphere in $\HE$ that contains $x$.
	Clearly, we may choose $\delta > 0$ so small that $B_{\delta}(x) = \{y \in \R^{d+1} \mid \Vert x - y \Vert < \delta\} \cap \HE$ is a topological ball contained in this sphere and contained in $V$. 
	The ball $B_\delta(x)$ can be contracted to $\{x\}$ in $V$, so choosing $U = B_{\delta}(x)$, we obtain that the inclusion $f_{\leq s} \cap U \to f_{\leq t} \cap V$ is homotopic to the constant map with value $x$.

@@

Both singular and \v{C}ech homology of the $d$-dimensional Hawaiian earring is infinite dimensional in degree $d$, so they satisfy the assumptions of \cref{t:counterexample}.
For singular homology this is proven in \cite{Barratt.1962}.
For \v{C}ech homology, one can use the fact that it commutes with totally ordered limits for compact Hausdorff spaces \cite[Theorems VIII.3.6 and X.3.1]{Eilenberg.1952} as follows.
Define
\begin{align*}
\HE_k &=
\left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle| \ \left( x_0 - \frac{1}{k} \right)^2 + x_1^2 + \dots + x_d^2 \leq \left( \frac{1}{k} \right)^2 \right\} \\ &\, \cup
\bigcup_{n=1}^{k-1} \left\{ (x_0, \dots, x_d) \in \R^{d+1} \ \middle |\ \left( x_0 - \frac{1}{n} \right)^2 + x_1^2 + \dots + x_d^2 = \left( \frac{1}{n} \right)^2 \right\},
\end{align*}
i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
We have $\lim_{k} \HE_{k} = \bigcap_{k} \HE_{k} = \HE$, and hence $\CH_{d}(\HE; \F) = \lim_{k} \CH_{d}(\HE_{k}; \F)$, where $\CH$ denotes \v{C}ech homology.
Clearly, each $\HE_{k}$ is a CW-complex, so we can simply use cellular homology to compute
\begin{equation*}
\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k}; \F)=\lim\left(\dots\to \prod_{n=1}^2\F\to \prod_{n=1}^1\F\to \prod_{n=1}^0\F\right)=\prod_{n\in\N}\F,
\end{equation*}
which is infinite-dimensional over $\F$.

\begin{cor} \label{c:counterexample}
	The function $f \colon \HE \to \R$ with value~$0$ at the origin and $1$ elsewhere defines a weakly $\LC$ compact sublevel set filtration that is not q-tame with respect to singular and \v{C}ech homology.
\end{cor}

### 2021-09-13 sec/vietoris.tex
Return to original paper size and use biblatex for references
Sent to CCM

First, let us recall the definition of \v{C}ech homology as presented for example by \citet[Section~IX--X]{Eilenberg.1952}.
Let $X$ be a topological space and let $\Cov(X)$ be the set of all open covers of $X$ ordered by the refinement relation. 
Recall that for an open cover $\alpha \in \Cov(X)$ its \emph{nerve} $\Nrv(\alpha)$ is defined as the simplicial complex
\begin{equation*}
\Nrv(\alpha) =
\big\{ \beta \subseteq \alpha \mid \beta \text{ is finite and } \textstyle{\bigcap_{U \in \beta}} \, U \neq \emptyset \big\}.
\end{equation*}
The nerve construction defines a functor from the poset $\Cov(X)$ regarded as a category to that of simplicial complexes. 
The \emph{\v{C}ech homology with coefficients in $\F$} of $X$ is defined as
\begin{equation*}
\CH(X; \F) \ =
\lim_{\alpha \in \Cov(X)} H(\Nrv(\alpha); \F).
\end{equation*}
As an alternative to the nerve construction, for a cover $\alpha \in \Cov(X)$ one can define $\Vietoris(\alpha)$ as the simplicial complex
\begin{equation*}
\Vietoris (\alpha) = \left\{ \sigma \subseteq X \mid \sigma \text{ is finite and } \sigma \in U \text{ for some } U \in \alpha \right\},
\end{equation*}
which again yields a functor from $\Cov(X)$ to simplicial complexes.
This construction is dual to the nerve construction in the sense of Dowker's Theorem \cite{Dowker.1952}, which asserts that the two complexes $\Nrv (\alpha)$ and $\Vietoris (\alpha)$ are homotopy equivalent.
As a consequence, we have that $H (\Nrv (\alpha); \F) \cong H (\Vietoris (\alpha); \F)$.
This isomorphism is natural with respect to refinement of covers, so we get an alternative description of \v{C}ech homology as 
\begin{equation*}
\CH (X; \F) \ \cong
\lim_{\alpha \in \Cov (X)} H (\Vietoris (\alpha); \F).
\end{equation*}

If $X$ is a metric space, this is still not exactly the same as the construction of metric Vietoris homology presented in \cite{Vietoris.1927} and used by Morse, which in modern notation is the limit
\begin{equation*}
\lim_{\alpha \in \Balls(X)} H (\Vietoris (\alpha); \F),
\end{equation*}
where 
\begin{equation*}
\Balls (X) = \left\{ ( B_{\delta} (x) )_{x \in X} \mid \delta > 0 \right\}
\subseteq \Cov (X).
\end{equation*}
However, if the metric space $X$ is compact, then $\Balls (X)$ is coinitial in $\Cov (X)$, that is to say, they both define the same limit.
Thus, we have a natural isomorphism
\begin{equation*}
\CH (X; \F) \ \cong \,
\lim_{\alpha \in \Balls(X)} H (\Vietoris (\alpha); \F)
\end{equation*}
for a compact metric space $X$.
In other words, the metric Vietoris homology theory employed in Morse's setting is canonically isomorphic to the \v{C}ech homology we consider in this paper.

>>>>

First, let us recall the definition of \v{C}ech homology as presented for example by \cite[Section~IX--X]{Eilenberg.1952}.
Let $X$ be a topological space and let $\Cov(X)$ be the set of all open covers of $X$ ordered by the refinement relation, which we implicitly regard as a poset category.
Recall that for an open cover $\alpha \in \Cov(X)$ its \emph{nerve} $\Nrv(\alpha)$ is defined as the simplicial complex
\begin{equation*}
\Nrv(\alpha) =
\big\{ \beta \subseteq \alpha \mid \beta \text{ is finite and } \textstyle{\bigcap_{U \in \beta}} \, U \neq \emptyset \big\}.
\end{equation*}
For any field $\F$, the nerve construction composed with the functor of simplicial homology with coefficients in $\F$ define a functor from $\Cov(X)$ to the category of graded $\F$-vector spaces.
The \emph{\v{C}ech homology with coefficients in $\F$} of $X$ is defined as
\begin{equation*}
\CH(X; \F) \ =
\lim_{\alpha \in \Cov(X)} H(\Nrv(\alpha); \F).
\end{equation*}
As an alternative to the nerve construction, for a cover $\alpha \in \Cov(X)$ one can define $\Vietoris(\alpha)$ as the simplicial complex
\begin{equation*}
\Vietoris (\alpha) = \left\{ \sigma \subseteq X \mid \sigma \text{ is finite and } \sigma \in U \text{ for some } U \in \alpha \right\}.
\end{equation*}
This construction is dual to the nerve construction in the sense of Dowker's Theorem \cite{Dowker.1952}, which asserts that the two complexes $\Nrv (\alpha)$ and $\Vietoris (\alpha)$ are homotopy equivalent after geometric realization.
As a consequence, we have that $H (\Nrv (\alpha); \F) \cong H (\Vietoris (\alpha); \F)$.
This isomorphism is natural with respect to refinement of covers, so we get an alternative description of \v{C}ech homology as
\begin{equation*}
\CH (X; \F) \ \cong
\lim_{\alpha \in \Cov (X)} H (\Vietoris (\alpha); \F).
\end{equation*}

If $X$ is an arbitrary metric space, this is still not exactly the same as the construction of \emph{metric} Vietoris homology, as originally defined by Vietoris \cite{Vietoris.1927} and used by Morse, which in modern notation is the limit over all covers of $X$ by metric $\delta$-balls,
\begin{equation*}
\lim_{\alpha \in \Balls(X)} H (\Vietoris (\alpha); \F),
\end{equation*}
where
\begin{equation*}
\Balls (X) = \left\{ ( B_{\delta} (x) )_{x \in X} \mid \delta > 0 \right\}
\subseteq \Cov (X).
\end{equation*}
However, if the metric space $X$ is compact, then $\Balls (X)$ is coinitial in $\Cov (X)$, that is to say, they both define the same limit.
Thus, in this case we have a natural isomorphism
\begin{equation*}
\CH (X; \F) \ \cong \,
\lim_{\alpha \in \Balls(X)} H (\Vietoris (\alpha); \F)
\end{equation*}
In other words, for compact metric spaces the metric Vietoris homology theory employed in Morse's setting is canonically isomorphic to the \v{C}ech homology we consider in this paper.
